---
title: "Sensitivity analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sensitivity analysis-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
number-sections: true
number-equations: true
crossref:
  chapters: true
link-citations: true
---

> *Work in progress!* 🏗

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, include = FALSE}
library(mcmodule)
```

# Risk analysis sensitivity analysis

Definition, correlation, Morris and Sobol indices.

Other packages needed

```{r, warning=FALSE, message=FALSE}
library(sensitivity)
library(dplyr)
library(ggplot2)
```

# Correlation analysis

The `mcmodule_corr()` function calculates correlation coefficients between input and output nodes across your mcmodule. This helps identify which input parameters have the strongest influence on your results. It is based on `mc2d::tornado()` but applied to many distinct variates.

```{r}
# Calculate correlations using Spearman's method (default)
cor_results <- mcmodule_corr(imports_mcmodule)

# View correlation results
head(cor_results)

# Use different correlation method
cor_pearson <- mcmodule_corr(imports_mcmodule, method = "pearson")
cor_kendall <- mcmodule_corr(imports_mcmodule, method = "kendall")
```

Visualize results

```{r}
# Plot tornado plot of correlations
mcmodule_tornado(imports_mcmodule)
```


# Morris index

```{r}
# Get bounds for Morris sampling design
b <- mctable_bounds(imports_mctable, transformation = FALSE)

sa <- sensitivity::morris(
  model = NULL,
  factors = b$factors,
  r = 10,
  design = list(type = "oat", levels = 4, grid.jump = 2),
  binf = b$binf,
  bsup = b$bsup,
  scale = TRUE
 )

imports_mcmodule_morris<-eval_module(
  exp = imports_exp,
  sample_design = sa,
  mctable = imports_mctable
)

imports_mcmodule_morris<-trial_totals(
  imports_mcmodule_morris,
  mc_names = c("no_detect_a"),
  trials_n = "animals_n",
  subsets_n = "farms_n",
  subsets_p = "h_prev",
  sample_design = sa,
  mctable = imports_mctable
)

y <-  unmc(imports_mcmodule_morris$node_list$no_detect_a_set$mcnode)

sensitivity::tell(sa, y)

sa
```


# Sobol index using `sensobol` package

```{r}
N = 10000
X <- mctable_sobol_matrices(imports_mctable, N = N)
imports_mcmodule_sobol <- eval_module(
  exp = imports_exp,
  data = NULL,
  sample_design = X,
  mctable = imports_mctable
)
imports_mcmodule_sobol <- trial_totals(
  mcmodule = imports_mcmodule_sobol,
  mc_names = c("no_detect_a"),
  trials_n = "animals_n",
  subsets_n = "farms_n",
  subsets_p = "h_prev",
  sample_design = X,
  mctable = imports_mctable
)

y <- unmc(imports_mcmodule_sobol$node_list$no_detect_a_set$mcnode)

## Compute and bootstrap the Sobol' indices:
sobol_result <- sobol_indices(Y = y, N = N, params = colnames(X))

sobol_result
```


