---
title: "Sensitivity analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sensitivity analysis-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
number-sections: true
number-equations: true
crossref:
  chapters: true
link-citations: true
---

> *Work in progress!* 🏗

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, include = FALSE}
library(mcmodule)
```

# Risk analysis sensitivity analysis

Definition, correlation, Morris and Sobol indices.

Other packages needed

```{r, warning=FALSE, message=FALSE}
library(sensitivity)
library(dplyr)
library(ggplot2)
```

# Correlation analysis

The `mcmodule_corr()` function calculates correlation coefficients between input and output nodes across your mcmodule. This helps identify which input parameters have the strongest influence on your results. It is based on `mc2d::tornado()` but applied to many distinct variates.

```{r}
# Calculate correlations using Spearman's method (default)
cor_results <- mcmodule_corr(imports_mcmodule)

# View correlation results
head(cor_results)

# Use different correlation method
cor_pearson <- mcmodule_corr(imports_mcmodule, method = "pearson")
cor_kendall <- mcmodule_corr(imports_mcmodule, method = "kendall")
```

Visualize results

```{r}
# Plot tornado plot of correlations
mcmodule_tornado(imports_mcmodule)
```


# Morris index

Get sampling design matrix

```{r}
eval_module(
  exp = imports_exp,
  data = imports_data,
  mctable = imports_mctable
)

a <- eval_module(
  exp = imports_exp,
  data = imports_data,
  mctable = imports_mctable
)

X1 <- sample_design(
  imports_mctable, 
  transformation = FALSE, 
  method = "morris",
  mc_names = c("h_prev", "w_prev", "test_sensi", "farms_n", "animals_n", "test_origin"),
  if_not_sample = "median")

b <- eval_module(
  exp = imports_exp,
  data = NULL,
  mctable = imports_mctable,
  sample_design = X1
)


X2 <- as.matrix(b$node_list$no_detect_a$mcnode)
#mcnode_to_matrix()

b <- trial_totals(
  mcmodule = b,
  mc_names = "no_detect_a",
  trials_n = "animals_n",
  subsets_n = "farms_n",
  subsets_p = "h_prev",
  mctable = imports_mctable,
)

```


# Sobol index


