---
title: "Sensitivity analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sensitivity analysis-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
number-sections: true
number-equations: true
crossref:
  chapters: true
link-citations: true
---

> *Work in progress*

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, include = FALSE}
library(mcmodule)
```

# Introduction

In risk analysis, sensitivity analysis answers a practical question: *which uncertain inputs matter most for the decision-relevant output?*

This vignette shows how to do that in `mcmodule` using three complementary approaches:

1. Correlation (tornado) analysis on an existing model run.
2. Morris elementary effects for screening.
3. Sobol indices for variance decomposition.

The goal is not to force one method to be "the truth". The goal is to build a stable understanding of which inputs drive risk, and why.
These methods answer different questions; use them together to build a robust picture of which inputs drive risk and why.

## Additional packages used

```{r, warning=FALSE, message=FALSE}
library(sensitivity)
library(sensobol)
library(dplyr)
library(ggplot2)
```

# Correlation analysis

`mcmodule_corr()` computes correlation coefficients between input and output nodes across the module. It gives a fast, first ranking of influential inputs and is based on `mc2d::tornado()`, extended to multiple variates.

It computes correlations across the rows of the module runs, using the uncertain input nodes and the chosen output node.

Unlike methods that require a sample design, this approach works directly on an `mcmodule` based on model data and does not require a dedicated `sample_design`. `mcmodule_corr()` computes correlations for uncertain inputs (stochastic nodes built with `func`) and, when `variates_as_nsv = TRUE`, also for multi-variate inputs.

First, compute correlations (Spearman by default) and inspect the table.

```{r}
# Calculate correlations using Spearman's method (default)
cor_results <- mcmodule_corr(imports_mcmodule, output = "no_detect_a", method = "spearman")

# View correlation results
head(cor_results)
```

Then visualise the ranking with a tornado plot. The output is a ggplot object, so you can customise it with standard `ggplot2` layers.

```{r}
# Plot tornado plot of correlations
mcmodule_tornado(imports_mcmodule, output = "no_detect_a", method = "spearman") + 
  labs(title = "Tornado Plot of Correlations") +
  theme(axis.text.y = element_text(size = 10))
```

In this example, `w_prev` is the dominant driver of `no_detect_a` (very strong positive association), while `test_sensi` shows a weaker negative association.

For a risk analyst, this is the quickest way to identify where improved data quality may reduce decision uncertainty.

# Sensitivity analysis with sample design

## Create a sample design

Before building anything, let’s clarify two concepts that are easy to confuse:

- **Sample space**: the allowable range or distribution definition for each input from which those combinations are generated.

- **Sample design**: the specific set of sampled input combinations used to run the model (for example, a matrix with one row per scenario).

In short, the sample space defines *where* values can come from; the sample design defines *which* values are actually used in model runs.

Here we use a global design so the same sampling matrix is available to the model evaluation and to any downstream sensitivity functions. Passing the design directly is just as valid; the global option keeps the example compact.

A `sample_design` in `mcmodule` can be either:

- a matrix/data frame, or
- a sensitivity object from the `sensitivity` package (for example, a Morris design object).

The sampling matrix must have one column per input node and one row per sample. Each row is one scenario in the `mcnode` uncertainty dimension. When you provide `sample_design`, that uncertainty dimension is no longer sampled from the original model inputs; it is mapped to the values in your design. This is useful when you want to explore uncertainty for inputs that were fixed in the original data (for example, `farms_n` and `animals_n` below). Currently, `sample_design` supports one variate.

You can pass the sampling matrix (or sensitivity object) directly to `eval_module()`, or set it as a global default with `set_sample_design()`.

```{r, eval=FALSE}
# Create a sample design as a data frame
X <- data.frame(
  w_prev = runif(1000, 0.15, 0.6),
  h_prev = runif(1000, 0.02, 0.7),
  test_sensi = runif(1000, 0.8, 0.91),
  animals_n = sample(5:10, 1000, replace = TRUE),
  farms_n = sample(82:176, 1000, replace = TRUE),
  test_origin = runif(1000, 0, 1)
)

dim(X)
head(X)

# Set global sample design
set_sample_design(X)

# Retrieve current global sample design
current_design <- set_sample_design()

# Evaluate the module with the global sample design
imports_mcmodule_X <- eval_module(
  exp = imports_exp
)

imports_mcmodule_X <- trial_totals(
  imports_mcmodule_X,
  mc_names = c("no_detect_a"),
  trials_n = "animals_n",
  subsets_n = "farms_n",
  subsets_p = "h_prev"
)

mcmodule_tornado(imports_mcmodule_X, output = "no_detect_a_set", method = "spearman")

# Reset global sample design
reset_sample_design()
```

This 1000-sample design produces a different ranking from the first tornado analysis. Here, `h_prev` is the main positive driver of `no_detect_a_set`. `test_origin` is the next most influential input and is negatively associated with the output. `w_prev` and `animals_n` have smaller positive effects, while `farms_n` and `test_sensi` are weak or negligible within these bounds.

This change is expected: the analysis now reflects your designed sampling space rather than the original model inputs.

## Build an `mctable` for sample-design sensitivity analysis

For sample-design based methods, `mctable` defines how inputs are sampled. In particular, `mctable$sample_space` stores the bounds or distribution parameters used to generate sensitivity samples.

We use `mctable` here to extract bounds and to build designs that stay consistent with the distributions used in the model.

```{r}
imports_mctable[,c("mcnode","mc_func", "transformation","sample_space")]
```

In this table:

- `mcnode` names the model input node.
- `mc_func` defines the sampling distribution (for example `runif`, `rnorm`, `rpert`).
- `sample_space` provides distribution parameters (for example `min/max`, `mean/sd`, or `min/mode/max`).
- `transformation` can map sampled values before evaluation.

Two helper functions support common workflows from `mctable`:

- `mctable_bounds()` for bound extraction (for Morris and related approaches).
- `mctable_sobol_matrices()` for Sobol design matrices.

## Morris elementary effects

Morris is a screening method. It identifies influential inputs at moderate computational cost and highlights potential nonlinearity or interactions.

The arguments `r`, `levels`, and `grid.jump` control the number of trajectories, the number of levels per factor, and the step size between levels.

Start by deriving parameter bounds from `mctable`.

```{r}
# Get bounds for Morris sampling design
b <- mctable_bounds(imports_mctable, transformation = FALSE)
```

Build the Morris design with `sensitivity::morris()`. The resulting object can be passed directly as `sample_design`.

```{r}
morris_sa <- sensitivity::morris(
  model = NULL,
  factors = b$factors,
  r = 10,
  design = list(type = "oat", levels = 4, grid.jump = 2),
  binf = b$binf,
  bsup = b$bsup,
  scale = TRUE
)
```

Evaluate the module with that design.

```{r}
imports_mcmodule_morris <- eval_module(
  exp = imports_exp,
  sample_design = morris_sa,
  mctable = imports_mctable
)
```

Compute the target output and aggregate where needed.

```{r}
imports_mcmodule_morris <- trial_totals(
  imports_mcmodule_morris,
  mc_names = c("no_detect_a"),
  trials_n = "animals_n",
  subsets_n = "farms_n",
  subsets_p = "h_prev",
  sample_design = morris_sa,
  mctable = imports_mctable
)
```

Extract the output vector and estimate Morris indices.

```{r}
y <- unmc(imports_mcmodule_morris$node_list$no_detect_a_set$mcnode)

# Aggregated output used for sensitivity analysis.
sensitivity::tell(morris_sa, y)

morris_sa
```

```{r}
plot(morris_sa)
```

In this example, `h_prev` has by far the largest `mu.star` (overall influence magnitude) and a high `sigma` (nonlinearity and/or interaction effects). This indicates that `h_prev` is both influential and not purely additive. `farms_n` is a secondary contributor, while `w_prev`, `test_sensi`, `animals_n`, and `test_origin` have much smaller elementary effects.


## Sobol indices

Sobol analysis decomposes output variance into first-order and higher-order effects.

First, generate Sobol sampling matrices from `mctable`.

`mctable_sobol_matrices()` creates Sobol design matrices from your `mctable` definitions.

```{r}
N <- 1000
X <- mctable_sobol_matrices(imports_mctable, N = N)
```

Evaluate the module and compute the target output.

```{r}
imports_mcmodule_sobol <- eval_module(
  exp = imports_exp,
  data = NULL,
  sample_design = X,
  mctable = imports_mctable
)

imports_mcmodule_sobol <- trial_totals(
  mcmodule = imports_mcmodule_sobol,
  mc_names = c("no_detect_a"),
  trials_n = "animals_n",
  subsets_n = "farms_n",
  subsets_p = "h_prev",
  sample_design = X,
  mctable = imports_mctable
)
```

Then compute first-order (`Si`) and total-order (`Ti`) Sobol indices.

```{r}
y <- unmc(imports_mcmodule_sobol$node_list$no_detect_a_set$mcnode)

# Compute Sobol indices
sobol_sa <- sensobol::sobol_indices(Y = y, N = N, params = colnames(X))

sobol_sa
```

In this run, `h_prev` explains most of the output variance (`Si` around 0.96 and `Ti` around 0.98), while `farms_n` contributes a small but non-negligible share. The remaining parameters are near zero. Slightly negative `Si` estimates are interpreted as Monte Carlo noise, not true negative variance contributions.

The gap between `Ti` and `Si` suggests interactions and/or non-additivity, but that difference can be noisy when `N` is small.

# Comparing methods and interpreting differences

For risk decisions, interpretation matters as much as computation. The three methods answer related, but different, questions.

- **Correlation/tornado**: Which inputs move with the output in the current model runs?
- **Morris**: Which inputs have large local effects across the sampled space, and where do interactions/nonlinearity appear?
- **Sobol**: Which inputs explain output variance globally, including interaction effects?

In this vignette, the three methods agree that `h_prev` is highly influential once analysis is performed on a designed sample space. Secondary rankings and magnitudes vary. That is expected, not a contradiction.

The main reasons are:

1. **Different estimands**  
  Correlation measures association, Morris measures elementary effects, and Sobol measures variance contributions.

2. **Different sampling spaces**  
  The first tornado analysis uses the original model structure, while Morris/Sobol use ranges defined in `mctable$sample_space`.

3. **Different output definitions**  
  You compare `no_detect_a` in the initial correlation analysis and `no_detect_a_set`/`no_detect_a_set` after aggregation with `trial_totals()`. Aggregation can change influence patterns.

4. **Model nonlinearity and interactions**  
  Morris `sigma` and the Sobol gap between `Ti` and `Si` indicate behaviour that simple rank correlation may understate.

5. **Monte Carlo error and design size**  
  Finite sample size introduces estimation noise, especially for small effects.


