---
title: "Sensitivity analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sensitivity analysis-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
number-sections: true
number-equations: true
crossref:
  chapters: true
link-citations: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, include = FALSE}
#library(mcmodule)
```

Sensitivity analysis quantifies how changes in an input affect the model output. In risk analysis, it answers a practical question: *which uncertain inputs are more important for a decision-relevant output?* Answering this helps to understand the model behaviour even without real data, to identify where to intervene to reduce risk, and to prioritise reducing uncertainty in the inputs that matter most.

In `mcmodule`, you can perform sensitivity analysis using different methods:

-   **Scenario (“what-if”) analysis:** Compares output changes under targeted input variations, practical for model runs using real data, but does not produce formal sensitivity indices.
-   **Correlation (tornado) analysis:** Ranking of influential inputs by correlating each input uncertainty with output uncertainty across existing runs. Only captures inputs that are uncertain in those runs.
-   **Sample design based sensitivity analysis:** Evaluates the model on a predefined input design (e.g., Morris/Sobol) to estimate sensitivity over a chosen input space using packages such as **sensitivity** or **sensobol**. In `mcmodule`, this method is currently limited to one variate.

# Scenario analysis

Scenario analysis (or “what-if analysis”) is a non-statistical but still useful approach to assess input sensitivity. It does not produce sensitivity indices, but it will show how the output changes when you make specific changes to the inputs. This makes it a practical way to compare targeted input changes using real input data. It is especially helpful for complex models where the most important inputs depend on context (for example, when estimating disease-entry risk across countries, where certain pathways will be more important in some countries than in others). Scenario analysis can therefore help you identify the most sensitive inputs for a particular use case. This topic is covered in the [working with what-if scenarios](https://nataliaciria.com/mcmodule/articles/mcmodule.html#working-with-what-if-scenarios) section.

# Correlation analysis

`mcmodule_corr()` ranks inputs by correlating each input node with the chosen output node across the model runs. It extends `mc2d::tornado()` to work with `mcmodule` objects and multiple variates.

Unlike methods that require a sample design, this approach works directly on an `mcmodule` built from model input data. `mcmodule_corr()` computes correlations for uncertain inputs (stochastic nodes built with `func`) and, when `variates_as_nsv = TRUE`, also for multi-variate inputs.

First, compute correlations (Spearman by default) and inspect the table.

```{r}
# Calculate output (expected number of animals not detected)
imports_mcmodule_corr <- trial_totals(
  imports_mcmodule,
  mc_names = c("no_detect_a"),
  trials_n = "animals_n",
  subsets_n = "farms_n",
  subsets_p = "h_prev",
  mctable = imports_mctable
)

# Calculate correlations using Spearman's method (default)
corr_results <- mcmodule_corr(imports_mcmodule_corr, output = "no_detect_a_set_n", method = "spearman")
```

In this example, `h_prev` is the main driver of `no_detect_a_set_n` (very strong positive association). `w_prev` and `animals_n` also show positive correlation, while `test_sensi` is close to zero or slightly negative across variates.

```{r}
# View correlation results
head(corr_results)
```

Then visualise the ranking with a tornado plot. `ggplot2` must be installed to create the plot, and it can be customised with standard `ggplot2` layers.

```{r}
library(ggplot2)

# Plot tornado plot of correlations
mcmodule_tornado(imports_mcmodule_corr, output = "no_detect_a_set_n", method = "spearman", print_summary = FALSE)
```

Each row in the plot represents one input node. The coloured point marks the variate with the largest absolute correlation for that input, and that value is used to rank the inputs. The black point shows the median correlation across variates, which is often a useful summary when correlations vary across the model. The black horizontal line shows the full range of correlations across variates, so you can see whether an input is consistently influential or only influential in certain variates.

Note that `farms_n` and `test_origin` are not shown in this analysis because they are fixed in the original model data and therefore have zero correlation with the output. This is a key limitation of correlation-based methods: they can only identify influential inputs that vary in the original model runs. To explore uncertainty for fixed inputs, you need a method that evaluates a sample design.

# Sensitivity analysis with sample design

## Create a sample design

An input **sample space** is the allowable range of values it can take (or distribution definition). A **sample design** is a specific set of sampled input combinations used to run the model. The sample space defines *where* values can come from, while the sample design defines *which* values are actually used in model runs.

A `sample_design` in `mcmodule` can be either:

-   a matrix/data frame, or
-   a sensitivity object from the `sensitivity::sensitivity` package (for example, a Morris design object).

The **sampling matrix** must have **one column per input node and one row per sample**. Each row is one scenario in the mcnode uncertainty dimension. When you provide a sample_design, that uncertainty dimension is no longer sampled from the original model inputs; it is mapped to the values in your design. Currently, sample_design supports one variate.

You can pass the sampling matrix (or sensitivity object) directly to `eval_module()`, or set it as a global default with `set_sample_design()`. Using a global design ensures the same sampling matrix is available across all model steps, which is convenient when you run sensitivity analysis after building a complex model. You can also pass the design to each function call, but that is more verbose, and you must remove it later to run the model with the original data inputs.

```{r, eval=FALSE}
# Create a sample design as a data frame
X <- data.frame(
  w_prev = runif(1000, 0.15, 0.6),
  h_prev = runif(1000, 0.02, 0.7),
  test_sensi = runif(1000, 0.8, 0.91),
  animals_n = sample(5:10, 1000, replace = TRUE),
  farms_n = sample(82:176, 1000, replace = TRUE),
  test_origin = runif(1000, 0, 1)
)

dim(X)
head(X)

# Set global sample design
set_sample_design(X)

# Retrieve current global sample design
current_design <- set_sample_design()

# Evaluate the module with the global sample design
imports_mcmodule_X <- eval_module(
  exp = imports_exp
)

# Calculate output (expected number of animals not detected)
imports_mcmodule_X <- trial_totals(
  imports_mcmodule_X,
  mc_names = c("no_detect_a"),
  trials_n = "animals_n",
  subsets_n = "farms_n",
  subsets_p = "h_prev"
)

mcmodule_tornado(imports_mcmodule_X, output = "no_detect_a_set_n", method = "spearman")

# Reset global sample design
reset_sample_design()
```

This sample design produces a different ranking from the first tornado analysis, because we're using the inputs sampling space rather than the original model inputs. Here, `h_prev` is the main positive driver of `no_detect_a_set_n`. `test_origin` is the next most influential input and is negatively associated with the output. `w_prev` and `animals_n` have smaller positive effects, while `farms_n` and `test_sensi` are weak or negligible within these bounds.

## Build an `mctable` for sample design sensitivity analysis

For methods that use a sample design, `mctable` defines how inputs are sampled. In particular, `mctable$sample_space` stores the bounds or distribution parameters used to generate sensitivity samples.

```{r}
imports_mctable[,c("mcnode","mc_func", "sample_space")]
```

In this table:

-   `mcnode` names the model input node.
-   `mc_func` defines the sampling distribution (for example `runif`, `rnorm`, `rpert`).
-   `sample_space` provides distribution parameters (for example `min/max`, `mean/sd`, or `min/mode/max`).

Two helper functions support common sampling designs from `mctable`:

-   `mctable_bounds()` for bound (*min* and *max*) extraction, for Morris and related approaches.
-   `mctable_sobol_matrices()` for Sobol design matrices.

Currently, `mcmodule` only provides direct conversion from `mctable` to Sobol sampling designs. However, you can use `mctable_bounds()` to create many other sample designs using different approaches. The only requirement is that your sample design matrix contains columns for all the input nodes in your `mcmodule` model.

## Morris elementary effects

Morris is a screening method. It identifies influential inputs at moderate computational cost and highlights potential nonlinearity or interactions.

Start by getting parameter bounds from `mctable`.

```{r}
# Get bounds for Morris sampling design
b <- mctable_bounds(imports_mctable, transformation = FALSE)
```

Build the Morris design with `sensitivity::morris()`. The resulting object can be passed directly as `sample_design`. The arguments `factors`, `r`, and `design` control the the factors (inputs) sampled, the number of repetitions of the design, and the step size between sampled values, respectively. The arguments `binf` and `bsup` provide the bounds for each input, and `scale = TRUE` scales the inputs to [0,1] for better comparability of Morris indices across inputs with different ranges.

To have stable Morris indices in large sample designs (with many inputs), you need to run the model with a large number of repetitions, which can be computationally intensive. Always check the sensitivity indices stability in diferent runs before interpreting them, and adjust `r` accordingly. 

```{r}
library(sensitivity)

# Create Morris design
morris_sa <- sensitivity::morris(
  model = NULL,
  factors = b$factors,
  r = 2000,
  design = list(type = "oat", levels = 4, grid.jump = 2),
  binf = b$binf,
  bsup = b$bsup,
  scale = TRUE
)

# Evaluate the module with that design.
imports_mcmodule_morris <- eval_module(
  exp = imports_exp,
  sample_design = morris_sa,
  mctable = imports_mctable
)

# Calculate output (expected number of animals not detected)
imports_mcmodule_morris <- trial_totals(
  imports_mcmodule_morris,
  mc_names = c("no_detect_a"),
  trials_n = "animals_n",
  subsets_n = "farms_n",
  subsets_p = "h_prev",
  sample_design = morris_sa,
  mctable = imports_mctable
)
```

Finally, extract the output vector and estimate Morris indices. The `tell()` function updates the Morris object with the model output, and then you can print and plot the indices. 

```{r}
# Extract the output vector and estimate Morris indices.
y <- unmc(imports_mcmodule_morris$node_list$no_detect_a_set_n$mcnode)

# Aggregated output used for sensitivity analysis.
sensitivity::tell(morris_sa, y)

# Print Morris indices
morris_sa

# Plot Morris indices (mu.star and sigma)
plot(morris_sa)
```

`h_prev` has the largest `mu.star`, making it the most influential input overall, followed by `test_origin`. `w_prev` is the next most important, while `farms_n` and `animals_n` have more moderate effects, and `test_sensi` is negligible within the explored bounds (note that the uncertainty is small, 0.875-0.91). The large `sigma` values suggest nonlinearity and/or interaction effects.

## Sobol indices

Sobol analysis decomposes output variance into first-order and higher-order effects, providing a global sensitivity analysis. It is more computationally intensive than Morris, but it gives a more complete picture of input importance, including interactions.

First, generate Sobol sampling matrices from `mctable`.

`mctable_sobol_matrices()` creates Sobol design matrices from your `mctable` definitions.

```{r}
library(sensobol)

N <- 1000
X <- mctable_sobol_matrices(imports_mctable, N = N)
```

Evaluate the module, compute the target output as before.

```{r}
imports_mcmodule_sobol <- eval_module(
  exp = imports_exp,
  data = NULL,
  sample_design = X,
  mctable = imports_mctable
)

imports_mcmodule_sobol <- trial_totals(
  mcmodule = imports_mcmodule_sobol,
  mc_names = c("no_detect_a"),
  trials_n = "animals_n",
  subsets_n = "farms_n",
  subsets_p = "h_prev",
  sample_design = X,
  mctable = imports_mctable
)
```

Then compute first-order (`Si`) and total-order (`Ti`) Sobol indices and print and plot the results. The `Si` index quantifies the contribution of each input to output variance, while the `Ti` index captures the total contribution of each input, including interactions.

```{r}
y <- unmc(imports_mcmodule_sobol$node_list$no_detect_a_set_n$mcnode)

# Compute Sobol indices
sobol_sa <- sensobol::sobol_indices(Y = y, N = N, params = colnames(X))

sobol_sa

plot(sobol_sa)
```

Here `h_prev` explains the largest share of output variance, `test_origin` is the next most important input, followed by `w_prev`. `farms_n` and `animals_n` have smaller but non-zero effects, while `test_sensi` is negligible. Further exploration could be done using `sensobol::plot_scatter()` and `sensobol::plot_multiscatter()`. Exploring interactions and nonlinearity would require more computationally intensive methods, such as second-order Sobol indices.

## Other methods

Other sensitivity analysis methods include regression based approaches (for example, using `sensitivity::fast99()`), cooperative game theory approaches (for example, using `iml::shapley()`), and others. These methods are not currently implemented in `mcmodule`, but you can use the sample design approach to implement them yourself by creating the appropriate sample design and then applying the method to the model outputs.