---
title: "Sensitivity analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sensitivity analysis-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
number-sections: true
number-equations: true
crossref:
  chapters: true
link-citations: true
---

> *Work in progress!* 🏗

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, include = FALSE}
library(mcmodule)
```

# Risk analysis sensitivity analysis

Definition, correlation, Morris and Sobol indices.

Other packages needed

```{r, warning=FALSE, message=FALSE}
library(sensitivity)
library(dplyr)
library(ggplot2)
```

# Correlation analysis

The `mcmodule_corr()` function calculates correlation coefficients between input and output nodes across your mcmodule. This helps identify which input parameters have the strongest influence on your results. It is based on `mc2d::tornado()` but applied to many distinct variates.

```{r}
# Calculate correlations using Spearman's method (default)
cor_results <- mcmodule_corr(imports_mcmodule)

# View correlation results
head(cor_results)

# Use different correlation method
cor_pearson <- mcmodule_corr(imports_mcmodule, method = "pearson")
cor_kendall <- mcmodule_corr(imports_mcmodule, method = "kendall")
```

Visualize results

```{r}
cor_results%>%
  ggplot(aes(x=value, y=input))+
  geom_point()
```


# Morris index


# Sobol index


