---
title: "Sensitivity analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sensitivity analysis-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
number-sections: true
number-equations: true
crossref:
  chapters: true
link-citations: true
---

> *Work in progress!* 🏗

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, include = FALSE}
library(mcmodule)
```

# Introduction

This vignette shows how to identify influential inputs in `mcmodule`.
It starts with correlation (tornado) analysis on an existing `mcmodel`, then moves to sample-design-based approaches (Morris and Sobol) built from `mctable$sample_space`.

## Additional packages used

```{r, warning=FALSE, message=FALSE}
library(sensitivity)
library(sensobol)
library(dplyr)
library(ggplot2)
```

# Correlation analysis

`mcmodule_corr()` computes correlation coefficients between input and output nodes across the module.
It provides a fast first ranking of influential parameters and is based on `mc2d::tornado()` generalized to multiple variates.

Unlike Morris and Sobol, this approach is applied on an `mcmodel` based on model data and does not require a dedicated `sample_design` object. This allows analysis across different variates directly from the model output structure. Correlations are computed for inputs that carry uncertainty (stochastic nodes built with `func`) and, when `variates_as_nsv = TRUE`, also for inputs with multiple variates.

First, compute correlations (Spearman by default) and inspect the table.

```{r}
# Calculate correlations using Spearman's method (default)
cor_results <- mcmodule_corr(imports_mcmodule, method = "spearman")

# View correlation results
head(cor_results)
```

Then visualize the ranking with a tornado plot.

```{r}
# Plot tornado plot of correlations
mcmodule_tornado(imports_mcmodule)
```

In this example, `w_prev` is consistently the dominant driver of `no_detect_a` (very strong positive association), while `test_sensi` shows a weaker negative association.

# Build an `mctable` for sample-design sensitivity analysis

For sample-design-based methods, `mctable` defines how inputs are sampled. In particular, `mctable$sample_space` stores the domain or distribution parameters used to generate sensitivity samples.

The example `imports_mctable` can be used as a template:

```{r}
imports_mctable
```

In this table:

- `mcnode` names the model input node.
- `mc_func` defines the sampling distribution (for example `runif`, `rnorm`, `rpert`).
- `sample_space` provides distribution parameters (for example `min/max`, `mean/sd`, or `min/mode/max`).
- `transformation` can map sampled values before evaluation.

`sample_design` in `mcmodule` is intentionally flexible: it can be a matrix/data frame or a sensitivity object with sampled values for the expression input nodes. You can pass it directly to `eval_module()`, or set a global default with `set_sample_design()`.

```{r, eval=FALSE}
# Set global sample design
set_sample_design(X)

# Retrieve current global sample design
current_design <- set_sample_design()

# Reset global sample design
reset_sample_design()
```

Two helper functions support common sensitivity workflows from `mctable`:

- `mctable_bounds()` for bound extraction (for Morris and related approaches).
- `mctable_sobol_matrices()` for Sobol design matrices.

# Morris elementary effects

Morris is a screening method that identifies influential factors at moderate computational cost.

Start by deriving parameter bounds from `mctable`.

```{r}
# Get bounds for Morris sampling design
b <- mctable_bounds(imports_mctable, transformation = FALSE)
```

Build the Morris design with `sensitivity::morris()`. The resulting object can be passed directly as `sample_design`.

```{r}
morris_sa <- sensitivity::morris(
  model = NULL,
  factors = b$factors,
  r = 10,
  design = list(type = "oat", levels = 4, grid.jump = 2),
  binf = b$binf,
  bsup = b$bsup,
  scale = TRUE
)
```

Evaluate the module with that design.

```{r}
imports_mcmodule_morris <- eval_module(
  exp = imports_exp,
  sample_design = morris_sa,
  mctable = imports_mctable
)
```

Compute the target output and aggregate where needed.

```{r}
imports_mcmodule_morris <- trial_totals(
  imports_mcmodule_morris,
  mc_names = c("no_detect_a"),
  trials_n = "animals_n",
  subsets_n = "farms_n",
  subsets_p = "h_prev",
  sample_design = morris_sa,
  mctable = imports_mctable
)
```

Extract the output vector and estimate Morris indices.

```{r}
y <- unmc(imports_mcmodule_morris$node_list$no_detect_a_set$mcnode)

sensitivity::tell(morris_sa, y)

morris_sa
```

```{r}
plot(morris_sa)
```

In this example, `h_prev` has by far the largest `mu.star` (overall influence magnitude) and also a high `sigma` (nonlinearity and/or interaction effects), indicating it is both influential and non-additive. `farms_n` is a secondary contributor, while `w_prev`, `test_sensi`, `animals_n`, and `test_origin` have much smaller elementary effects.


# Sobol indices

Sobol analysis decomposes output variance into first-order and higher-order effects.

First, generate Sobol sampling matrices from `mctable`.

`mctable_sobol_matrices()` creates a Sobol design from your `mctable` definitions.

```{r}
N <- 1000
X <- mctable_sobol_matrices(imports_mctable, N = N)
```

Evaluate the module and compute the target output.

```{r}
imports_mcmodule_sobol <- eval_module(
  exp = imports_exp,
  data = NULL,
  sample_design = X,
  mctable = imports_mctable
)

imports_mcmodule_sobol <- trial_totals(
  mcmodule = imports_mcmodule_sobol,
  mc_names = c("no_detect_a"),
  trials_n = "animals_n",
  subsets_n = "farms_n",
  subsets_p = "h_prev",
  sample_design = X,
  mctable = imports_mctable
)
```

Then compute first-order (`Si`) and total-order (`Ti`) Sobol indices.

```{r}
y <- unmc(imports_mcmodule_sobol$node_list$no_detect_a_set$mcnode)

# Compute Sobol indices
sobol_sa <- sensobol::sobol_indices(Y = y, N = N, params = colnames(X))

sobol_sa
```

In this run, `h_prev` explains almost all output variance (`Si` around 0.96 and `Ti` around 0.98), while `farms_n` contributes a small but non-negligible share. The remaining parameters are near zero, and slightly negative `Si` values are interpreted as Monte Carlo estimation noise rather than true negative variance contributions.


