---
title: "Sensitivity analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sensitivity analysis-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
number-sections: true
number-equations: true
crossref:
  chapters: true
link-citations: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, include = FALSE}
library(mcmodule)
```

Sensitivity analysis quantifies how changes in an input, within its plausible range, affect the model output. In risk analysis, sensitivity analysis usually answers a practical question: *which uncertain inputs matter most for a decision-relevant output?*  

In `mcmodule`, you can answer that question with different methods. The quickest option on an existing model run is correlation (tornado) analysis, using `mcmodule_corr()` or `mcmodule_tornado()`. `mcmodule` also supports design-based sensitivity analyses, where you evaluate the model on a predefined input sample design generated by packages such as **sensitivity** or **sensobol**.

# Correlation analysis

`mcmodule_corr()` ranks inputs by correlating each input node with the chosen output node across the model runs. It extends `mc2d::tornado()` to work with `mcmodule` objects and multiple variates.

Unlike methods that require a sample design, this approach works directly on an `mcmodule` built from model input data. `mcmodule_corr()` computes correlations for uncertain inputs (stochastic nodes built with `func`) and, when `variates_as_nsv = TRUE`, also for multi-variate inputs.

First, compute correlations (Spearman by default) and inspect the table.

```{r}
# Calculate output: Expected number of animals not detected
imports_mcmodule_corr <- trial_totals(
  imports_mcmodule,
  mc_names = c("no_detect_a"),
  trials_n = "animals_n",
  subsets_n = "farms_n",
  subsets_p = "h_prev",
  mctable = imports_mctable
)

# Calculate correlations using Spearman's method (default)
corr_results <- mcmodule_corr(imports_mcmodule_corr, output = "no_detect_a_set_n", method = "spearman")
```

In this example, `h_prev` is the main driver of `no_detect_a_set_n` (very strong positive association). `w_prev` and `animals_n` also show positive correlation, while `test_sensi` is close to zero or slightly negative across variates.

```{r}
# View correlation results
head(corr_results)
```

Then visualise the ranking with a tornado plot. `ggplot2` must be installed to create the plot, and it can be customised with standard `ggplot2` layers.

```{r}
library(ggplot2)

# Plot tornado plot of correlations
mcmodule_tornado(imports_mcmodule_corr, output = "no_detect_a_set_n", method = "spearman", print_summary = FALSE)
```

Each row in the plot represents one input node. The coloured point marks the variate with the largest absolute correlation for that input, and that value is used to rank the inputs. The black point shows the median correlation across variates, which is often a useful summary when correlations vary across the model. The black horizontal line shows the full range of correlations across variates, so you can see whether an input is consistently influential or only influential in certain variates.

Note that `farms_n` and `test_origin` are not shown in this analysis because they are fixed in the original model data and therefore have zero correlation with the output. This is a key limitation of correlation-based methods: they can only identify influential inputs that vary in the original model runs. To explore uncertainty for fixed inputs, you need a method that evaluates a sample design.

# Sensitivity analysis with sample design

## Create a sample design

Before building anything, let’s clarify two concepts:

- **Sample space**: allowable range or distribution definition for each input

- **Sample design**: specific set of sampled input combinations used to run the model

The sample space defines *where* values can come from, while the sample design defines *which* values are actually used in model runs.

A `sample_design` in `mcmodule` can be either:

- a matrix/data frame, or
- a sensitivity object from the `sensitivity::sensitivity` package (for example, a Morris design object).

The sampling matrix must have one column per input node and one row per sample. Each row is one scenario in the mcnode uncertainty dimension. When you provide a sample_design, that uncertainty dimension is no longer sampled from the original model inputs; it is mapped to the values in your design. Currently, sample_design supports one variate.

You can pass the sampling matrix (or sensitivity object) directly to `eval_module()`, or set it as a global default with `set_sample_design()`. Using a global design ensures the same sampling matrix is available across all model steps, which is convenient when you run sensitivity analysis after building a complex model. You can also pass the design to each function call, but that is more verbose, and you must remove it later to run the model with the original data inputs.

```{r, eval=FALSE}
# Create a sample design as a data frame
X <- data.frame(
  w_prev = runif(1000, 0.15, 0.6),
  h_prev = runif(1000, 0.02, 0.7),
  test_sensi = runif(1000, 0.8, 0.91),
  animals_n = sample(5:10, 1000, replace = TRUE),
  farms_n = sample(82:176, 1000, replace = TRUE),
  test_origin = runif(1000, 0, 1)
)

dim(X)
head(X)

# Set global sample design
set_sample_design(X)

# Retrieve current global sample design
current_design <- set_sample_design()

# Evaluate the module with the global sample design
imports_mcmodule_X <- eval_module(
  exp = imports_exp
)

# Calculate output (at least one animal from at least one farm is not detected)
imports_mcmodule_X <- trial_totals(
  imports_mcmodule_X,
  mc_names = c("no_detect_a"),
  trials_n = "animals_n",
  subsets_n = "farms_n",
  subsets_p = "h_prev"
)

mcmodule_tornado(imports_mcmodule_X, output = "no_detect_a_set_n", method = "spearman")

# Reset global sample design
reset_sample_design()
```

This sample design produces a different ranking from the first tornado analysis. Here, `h_prev` is the main positive driver of `no_detect_a_set_n`. `test_origin` is the next most influential input and is negatively associated with the output. `w_prev` and `animals_n` have smaller positive effects, while `farms_n` and `test_sensi` are weak or negligible within these bounds.

This change is expected: the analysis now reflects your designed sampling space rather than the original model inputs.

## Build an `mctable` for sample design sensitivity analysis

For methods that use a sample design, `mctable` defines how inputs are sampled. In particular, `mctable$sample_space` stores the bounds or distribution parameters used to generate sensitivity samples.

We use `mctable` here to extract bounds and to build designs that stay consistent with the distributions used in the model.

```{r}
imports_mctable[,c("mcnode","mc_func", "sample_space")]
```

In this table:

- `mcnode` names the model input node.
- `mc_func` defines the sampling distribution (for example `runif`, `rnorm`, `rpert`).
- `sample_space` provides distribution parameters (for example `min/max`, `mean/sd`, or `min/mode/max`).

Two helper functions support common sampling designs from `mctable`:

- `mctable_bounds()` for bound extraction (for Morris and related approaches).
- `mctable_sobol_matrices()` for Sobol design matrices.

Currently, `mcmodule` only provides direct conversion from `mctable` to Sobol sampling designs. However, you can use `mctable_bounds()` to create many other sample designs using different approaches. The only requirement is that your sample design matrix contains columns for all the input nodes in your `mcmodule` model.

## Morris elementary effects

Morris is a screening method. It identifies influential inputs at moderate computational cost and highlights potential nonlinearity or interactions.

The arguments `r`, `levels`, and `grid.jump` control the number of trajectories, the number of levels per factor, and the step size between levels.

Start by getting parameter bounds from `mctable`.

```{r}
# Get bounds for Morris sampling design
b <- mctable_bounds(imports_mctable, transformation = FALSE)
```

Build the Morris design with `sensitivity::morris()`. The resulting object can be passed directly as `sample_design`.

```{r}
library(sensitivity)

morris_sa <- sensitivity::morris(
  model = NULL,
  factors = b$factors,
  r = 10,
  design = list(type = "oat", levels = 4, grid.jump = 2),
  binf = b$binf,
  bsup = b$bsup,
  scale = TRUE
)
```

Evaluate the module with that design.

```{r}
imports_mcmodule_morris <- eval_module(
  exp = imports_exp,
  sample_design = morris_sa,
  mctable = imports_mctable
)
```

Compute the target output and aggregate where needed.

```{r}
imports_mcmodule_morris <- trial_totals(
  imports_mcmodule_morris,
  mc_names = c("no_detect_a"),
  trials_n = "animals_n",
  subsets_n = "farms_n",
  subsets_p = "h_prev",
  sample_design = morris_sa,
  mctable = imports_mctable
)
```

Extract the output vector and estimate Morris indices.

```{r}
y <- unmc(imports_mcmodule_morris$node_list$no_detect_a_set_n$mcnode)

# Aggregated output used for sensitivity analysis.
sensitivity::tell(morris_sa, y)

morris_sa
```

```{r}
plot(morris_sa)
```

In this run, `test_origin` has the largest `mu.star` by far, and it is strongly negative on average. `h_prev` is the next most influential input, followed by `farms_n` and `animals_n`. `w_prev` is smaller again, and `test_sensi` is negligible. The large `sigma` values for `test_origin` and `h_prev` indicate nonlinearity and/or interaction effects.


## Sobol indices

Sobol analysis decomposes output variance into first-order and higher-order effects.

First, generate Sobol sampling matrices from `mctable`.

`mctable_sobol_matrices()` creates Sobol design matrices from your `mctable` definitions.

```{r}
library(sensobol)

N <- 1000
X <- mctable_sobol_matrices(imports_mctable, N = N)
```

Evaluate the module and compute the target output.

```{r}
imports_mcmodule_sobol <- eval_module(
  exp = imports_exp,
  data = NULL,
  sample_design = X,
  mctable = imports_mctable
)

imports_mcmodule_sobol <- trial_totals(
  mcmodule = imports_mcmodule_sobol,
  mc_names = c("no_detect_a"),
  trials_n = "animals_n",
  subsets_n = "farms_n",
  subsets_p = "h_prev",
  sample_design = X,
  mctable = imports_mctable
)
```

Then compute first-order (`Si`) and total-order (`Ti`) Sobol indices.

```{r}
y <- unmc(imports_mcmodule_sobol$node_list$no_detect_a_set_n$mcnode)

# Compute Sobol indices
sobol_sa <- sensobol::sobol_indices(Y = y, N = N, params = colnames(X))

sobol_sa
```

In this run, `h_prev` explains the largest share of output variance (`Si` around 0.33 and `Ti` around 0.49). `test_origin` is the next most important input (`Si` around 0.23 and `Ti` around 0.34), followed by `w_prev` (`Si` around 0.14 and `Ti` around 0.23). `farms_n` and `animals_n` have smaller but non-zero effects, while `test_sensi` is negligible. Slightly negative `Si` estimates would still be Monte Carlo noise rather than true negative variance contributions.

The gap between `Ti` and `Si` suggests interactions and/or non-additivity, but that difference can be noisy when `N` is small.

# Comparing methods and interpreting differences

For risk decisions, interpretation matters as much as computation. The three methods answer related questions, but they do not estimate the same quantity.

- **Correlation/tornado**: Which inputs move with the output in the current model runs?
- **Morris**: Which inputs have large local effects across the sampled space, and where do interactions/nonlinearity appear?
- **Sobol**: Which inputs explain output variance globally, including interaction effects?

In this vignette, the methods agree that `h_prev` matters, but they disagree on the secondary story. Correlation puts `h_prev` first, with `w_prev` and `animals_n` also visible. Morris ranks `test_origin` highest by absolute elementary effect, with `h_prev` second and `farms_n` and `animals_n` still contributing. Sobol again places `h_prev`, `test_origin`, and `w_prev` at the top, but now in terms of variance contribution.
